---
title: "Running Trend Analyses on GCAM Data"
author: "Robert Link"
date: "2017-06-17"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `trend_analysis` function allows you to identify a selection of shared trend
types in your GCAM output and categorize the data by trend type.  For example,
if you were looking at population data by region you might have some regions
where population is gowing, some mostly stable, and so on.  

## Preparing the data

To run the trend analysis your data should be in long format.  We'll use the
Reference scenario population as an example.
```{r setup}
library(GCAManalysis)
data(population)
head(population)
```
There must be columns with the year and the data to be analyzed.  They can have
any names you like; you'll get a chance to specify the names when you run the
analysis.  Here they are called `year` and `population`.  Any remaining columns
are id columns; they define the groupings of the data into time series.  In this
case there is only one, `region`.  If your data has columns that you don't want
to use as groupings (e.g., maybe you have data down to the subsector, but you
only want to analyze by region and sector), then you will need to aggregate
appropriately before you run the analysis.

## Usage

Run the trend analysis and save the results.
```{r analysis, echo=-1}
set.seed(8675309)
tr <- trend_analysis(population, n=4, valuecol='population', yearcol='year')
```
The `n` argument determines the number of trend categories to find.  You might
have to experiment a bit to find the right number of categories for any
particular dataset.  A good rule of thumb is to use about one tenth the number
of time series.  The remaining two arguments indicate which columns hold the
years and values.  They can be omitted if the columns have their default names
of "year" and "value", respectively.

The return value contains two data frames.  The `trend` element has the trends
that were identified.  Each row has one data point for one of the prototype
trends.  The `trend.category` column identifies the trend category.  The column
with the years will always be called `year`, regardless of what it was called in
the original data.  The value column will be normalized such that the largest
value is 1.  It will be named "normalized.whatever", where "whatever" is the
name of the value column in the original data, so in this example,
`normalized.population`.  

The `categories` element has the original data, normalized as in the trend
prototypes.  The `trend.category` column indicates which trend category each
time series belonged to.

The return object has a couple of methods defined to make analysis more
convenient.  The `summary` method produces a table of category assignments and a
count of the number of time series assigned to each category.

```{r summary}
summary(tr)
```

```{r plot, fig.width=7, fig.height=6}
plot(tr)
```
